Material Culture Networks in Tonto Basin

### 2022-12-28 manuscript.Rmd
finished updating paper based on Chris Caseldine's comments

---
title: Material Culture Networks in Tonto Basin
output:
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
  rticles::tf_article: null
author:
  - name: Robert J. Bischoff 1
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Tonto Basin is located in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions and was primarily occupied during the Roosevelt phase (AD 1275-1325) and the Gila phase (AD 1325-1450). The Roosevelt phase is notable for the appearance of Salado pottery, which has been interpreted as evidence of Kayenta immigration into Tonto Basin and the beginning of the Salado phenomenon. Salado pottery production was widespread, with production often centered at the location of former immigrant enclaves. This study uses data from the Roosevelt Community Platform Mounds Study to analyze social relations through networks based on architecture, ceramics, projectile points, and site locations. The results show a strong correlation between projectile points and space, and differences in the ceramic and point networks suggest that different social processes, possibly relating to gender, influenced the structure of the networks. Sites in this study with roomblocks--associated with immigration--are highly central in the ceramic networks, but have low centrality in the point networks, indication differences in the networks and suggesting differences in the levels of social integration between the genders.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
---

```{r setup, echo = F}
out.format <- knitr::opts_knit$get("out.format")
options(knitr.table.format = "latex")
knitr::opts_chunk$set(
  echo = F,
  message = FALSE,
  warning = FALSE,
  dpi = 450,
  out.width = "100%"
)
```

```{r load, include=FALSE}
library(tidyverse,quietly = T)
library(magrittr)
library(Momocs,quietly = T)
library(kableExtra)
library(ggthemes, quietly = T)
library(here)
library(rio)
library(igraph)
library(ggnetwork)
library(vegan)
library(multinet)

@@

Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku]. This analysis uses the available information from the Roosevelt Community Platform Mounds Study (RPMS), the largest of the studies [@Rice1998-ku], and builds on prior research [@Clark2001-ck; @Hill2015-at; @Huntley2016-em; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Stark1998-mu; @Watts2013-ub] to combine data on architecture, ceramics, projectile points, and site locations to analyze the networks of social relations that existed in the past through networks based on these data. The primary question is to determine whether these data are correlated. Because spatial data is included, this would mean that near sites are more similar than distant sites in architecture, ceramics, and projectile points. A lack of correlation would indicate that factors other than distance are influencing the social networks. The results of the network analysis demonstrate a strong correlation between  projectile points and space. The differences in the ceramic and point networks indicate that different social processes, perhaps relating to gender, are affecting the structure of the network.

Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`). The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the Salt River. It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s. All of the artifacts/data from the RPMS project are hosted at Arizona State University, and the data in this study come exclusively from this project.

@@

Most of the occupation of the Tonto Basin, particularly the sites excavated during the project, date between AD 1275 and 1325, which corresponds to the Roosevelt Phase. The Roosevelt phase is notable for the appearance of Salado pottery. Moderate occupation continued into the Gila phase (AD 1325-1450), after which recognizable occupation ceased in the area until the historic period.

One aspect of Tonto Basin that has caught the attention of many researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin and the beginning of the Salado phenomenon [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu]. Salado pottery production was widespread, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].

Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity. Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study, however. His study of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin. He found strong connections between many parts of the eastern Tonto Basin.

One way to examine individuals who make similar types of projectile points--and similar types of pottery for that matter--is to talk about identity. Identity can be a troubling topic for anthropologists. There are numerous meanings given to it and numerous scales at which it applies [@Brubaker2000-na], but Peeples [-@Peeples2018-ib] has introduced archaeologists to a more practical approach that is useful here. This approach views identity as existing along two axes: one categorical and the other relational. Relational identification is the process of identifying with someone due to frequent interaction. Categorical identification is the process of identifying with someone because you belong to a recognized social group. For example, members of the same moiety would share a categorical identity. They would likely also share a relational identity if they frequently interacted. These identities can be reflected in material culture. Pottery makes a good example. A potter can produce large, black and red chevrons on a pot that everyone who sees it can associate with a particular social group. Members of that group may wish to use that pottery as well as those who wish to curry favor with that group. Others may avoid that pottery. This pottery would be a marker of categorical identity. Relational identity can be seen in the way the potter prepares the clay recipe and smooths the coils. The particulars of these actions would be learned from close interaction with a teacher or other potters. In this way, relational identity can be compared to communities of practice [@Lave1991-ik; @Wenger1998-rx].

In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity. The architecture discussed is highly visible and representative of historical group membership. The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership [see @Clark2001-ck for a discussion on material culture and its relationship to identity]. Projectile points can in some cases indicate group membership, but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww]. Styles were difficult to determine for the original researchers, and I had the same difficulty. In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case. There are no environmental reasons to assume differences in projectile points. Each hunter would have been using the points for the same game.

Another way that people identify themselves is gender. Gender roles, like identity, can also vary, but for the purposes of this analysis I will use a simplified model. Women made pottery and men made projectile points. This was not always true of course, but it generally fits the ethnographic record in the Southwest. This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but I will make the assumption that ceramic networks and point networks are gendered. Therefore differences in these networks represent different interactions between men's and women's networks.

@@

The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between sites [@Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

Network approaches are best suited where the relationships between entities are of primary interest. In this case, I compare the similarity between sites using architecture, ceramics, and projectile points. Thus, a network can be defined using a site from Tonto Basin as a node and a type of similarity as a link (or edge) between the nodes. Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd]. These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts.

Table `r tbls` 1 lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site. The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used. The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku].

```{r sites, echo=FALSE}
tbls = tbls + 1
cap = "Project Sites and Data"
sites %>% 
  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>% 
knitr::kable(caption= cap, 
             booktabs = T,
             format.args = list(big.mark = ",",
  scientific = FALSE)) %>% 
  kable_styling(latex_options = "scale_down")
```

The types of similarity used in this study are varied depending on the type of network. One parameter that is consistent is how similar links should be in order to keep them. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then cosider each link of equal value--has is drawbacks [@Peeples2013-tj], but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links.

I used ranked links to keep consistent values between networks. For example, a network could be made of the `r nrow(sites)` sites as nodes with each node being connected to the three sites that it has the strongest similarity to for ceramics, architecture, projectile points, and the three sites that it is physically closest to for the spatial network. In practice, not every node will have three connections. More than three ties can exist when the ties are tied in rank. Some nodes may not have three connections. And finally, because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top three of multiple other nodes. In the latter case, this is a good indication of a central node in the network. Because I am using several types of networks, there is no common cutoff value I can use. Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top three to ten ties. Examining the range of these metrics provides a more robust analysis.

Two network metrics are discussed in this analysis: eigenvector centrality and multilayer network correlation--the second will be discussed in the next section. Node centrality is a common way to quantify networks [@Borgatti2005-iz]. Centrality is a measure of the influence of the node on the network. Nodes with higher centrality generally derive or generate greater benefits from the network. Eigenvector centrality is one way to measure node centrality and is commonly used in archaeology. This metric describes how well a node is connected to the network as a whole and is helpful in comparing different networks containing the same nodes. Eigenvector centrality is also more robust to missing nodes than other measures [@Peeples2017-kl], a major problem in most archaeological studies.

## Multilayer Networks {-}

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network).

Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results.

@@

The simplest network is the spatial network. This network was created by calculating the Euclidean distance from every site to every other site. This method is equivalent to proximal point analysis, which was used in one of the earliest network studies in archaeology [@Terrell1977-co] and continues to be used to study potential pathways of interaction [see @Broodbank2000-pg; @Collar2013-fb].

@@

The architectural components consisted of compounds, roomblocks, and platform mounds. Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku]. Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture. Thus, if a site is labeled a platform mound it may have other architecture. The network for architecture, and for the remaining networks, was created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site. This is a simple measure of similarity that does not take into account abundance. It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.

@@

The ceramic network consists of 55 types--plainware and decorated--ranging from 181,666 Plain Brown Ware sherds to a single Black Mesa Black-on-white sherd. A single sherd is not much, but is still indicative of trade. However, the median sherd count was 157 for all types. The ceramic data was accessed from the cyberSW database [@Mills2020-tb], which keeps the full citations for the ceramic data.

@@

The point networks were expected to represent relational identification among men. This is not always the case with projectile points. In some cases they can represent group identity [see @Wiessner1983-ei], but in this case there is little variation in style. Because relational identification is related to frequent interaction, spatial distance is a crucial component. It is much harder to interact with someone when they are far away. Thus, the correlation between the point and spatial networks also makes theoretical sense.

>>>>

---
title: Material Culture Networks in Tonto Basin
output:
  rticles::tf_article: null
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
author:
  - name: Robert J. Bischoff
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type. Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities. This study takes architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data into a multilayer network analysis. This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation. The results demonstrate a strong association between projectile points and spatial distance. Sites containing roomblock architecture are highly central in the ceramic network but have low centrality in the projectile point network. Overall, the ceramic and projectile point networks exhibit significant differences. This indicates that the social networks that created these patterns had different social mechanisms. One potential cause of these differences is gendered spheres of interaction with men exchanging projectile points and women exchanging ceramics.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
always_allow_html: true
---

```{r setup, echo = F}
out.format <- knitr::opts_knit$get("out.format")
# options(knitr.table.format = "latex")
knitr::opts_chunk$set(
  echo = F,
  message = FALSE,
  warning = FALSE,
  dpi = 450,
  out.width = "100%"
)
```

```{r load, include=FALSE}
library(tidyverse,quietly = T)
library(magrittr)
library(Momocs,quietly = T)
library(kableExtra)
library(huxtable)
library(ggthemes, quietly = T)
library(here)
library(rio)
library(igraph)
library(ggnetwork)
library(vegan)
library(multinet)

@@

Network science has many applications for archaeologists. It can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb]. Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh]. Social systems are complex webs of interrelations. There is too much involved to model the material culture of even a technologically simple society. But certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand. Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.

Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture. For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large scale social networks. Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region. This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework. Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated. This is an important question because archaeologists often make inferences based solely on ceramics or other types of material culture. Evidence regarding the ways material culture co-vary will help archaeologists direct future research. Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.

Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub]. This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and a recent projectile point analysis [@Bischoff2022-rg]. There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites. In reality, I expected significant variation between these types of material culture. As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance. Furthermore, ceramic and projectile point networks (hereafter point networks) exhibit significant differences. I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.

## Tonto Basin

Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`). Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview]. The basin was also conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p. 129-133]. The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the confluence of the Salt River and Tonto Creek. It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s. Some important regional differences may have differentiated sites along the Salt River and Tonto Creek arms of Roosevelt Lake [@Lyons2013-ya; @Simon2001-am] based on non-local ceramics and obsidian. All of the artifacts/data from the RPMS project are hosted at the Center for Archaeology and Society at Arizona State University, and the data in this study come exclusively from this project.

@@

Tonto Basin was occupied, although likely not continuously, from the Archaic through the late Classic period. The height of occupation, particularly the sites excavated during the project, dates between AD 1275 and 1325, which corresponds to the Roosevelt Phase. The Roosevelt phase is notable for the appearance of Salado pottery and platform mounds--probably introduced from the Phoenix Basin [@Elson1996-ym]. Moderate occupation continued into the Gila phase (AD 1325-1450), which was characterized by population loss and aggregation into larger sites. This phase marks the end of recognizable occupation in the area until the historic period.

One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze]. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu]. This migration into the Tonto Basinwas part of a larger migration from north to south and is associated with the origin of the Salado phenomenon. Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd]. The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks. Having high centrality means they have more connections to other nodes (i.e., sites). The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between sites [e.g., @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

Network analysis is a flexible tool for analyzing many types of data [see @Brughmans2023-uj]. The key element is that some way must be determined to tie each entity together. This analysis uses sites as nodes in the network and some type of similarity as a tie to connect each node. Each type of similarity depends on the type of data. Table `r tbls` lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site. The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used. The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku]. This is not a random sample of Tonto Basin sites. It is merely a sample of convenience. Sampling can have a major impact on conclusions. In this case, I am not drawing conclusions based on a complete network or the existence of a random sample (though biases can still affect the results). Rather, I am interested in whether the types of material culture present in this study co-vary with each other and with space. These questions do not require a complete network.

## Defining Networks and Relations

Network approaches are best suited where the relationships between entities are of primary interest. In this case, the entities are defined as archaeology sites. The archaeology sites in turn represent a group of individuals who lived or visited the area of the site and left material remains behind. Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity. Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study. His analysis of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin. He found strong connections between many parts of the eastern Tonto Basin. Unfortunately, this study will only attempt a site-level analysis. Watt's study does, however, illustrate how relations can be defined between nodes in a network analysis. He used similarity networks where the points that had similar flaking styles were connected together. Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd]. These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts. This study also uses similarity networks to group sites by similarity in ceramic assemblages and similarity in projectile point forms.

Yet, it can be difficult to understand how a similarity network is related to a past social network. What kinds of interactions are represented by similarities in projectile points or in ceramics? Answering this question is also a crucial step in interpreting networks. One way to examine relations between individuals who make or use similar types of projectile points or pots is to talk about identity. Identity can be a troubling topic for anthropologists. There are numerous meanings given to it and numerous scales at which it applies [@Brubaker2000-na], but Peeples [-@Peeples2018-ib] has introduced archaeologists to a more practical approach that is useful here. This approach views identity as existing along two axes: one categorical and the other relational. Relational identification is the process of identifying with someone due to frequent interaction. Categorical identification is the process of identifying with someone because you belong to a recognized social group. For example, members of the same moiety would share a categorical identity. They would likely also share a relational identity if they frequently interacted. These identities can be reflected in material culture. Pottery makes a good example. Peeples [-@Peeples2018-ib, p. 151] notes that bowls in the Cibola region were sometimes painted with a bright, red slip with designs painted on the interior and they were among the first in the Southwest to have designs on the exterior. This is a strong indication of categorical identity. The potters were attempting to make a clear signal for whoever saw the pot. Relational identity can be seen in the way the potter prepares the clay recipe and smooths the coils. The particulars of these actions would be learned from close interaction with a teacher or other potters. In this way, relational identity can be compared to communities of practice [@Lave1991-ik; @Wenger1998-rx].

In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity. The architecture discussed is highly visible and representative of historical group membership. The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership. What this means is that links between nodes in the cer The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study. Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity. He identifies certain patterns, but his synthesis and other research [@Carr1995-mq; @Carr1995-sh; @Dietler1998-zo; @Gosselain1998-qd; @Gosselain2000-wf; @Gosselain2016-as; @Hodder1982-no; @Huntley2008-qu; @Lemonnier1986-le; @Lyons2003-yy; @Neuzil2008-zd; @Sassman2001-sw; @Stark1998-mu; @Wiessner1983-ei; @Wiessner1997-su] strongly indicates that the relationship between material culture and identity is culturally relative. Projectile points can in some cases indicate group membership [@Wiessner1983-ei; @Wiessner1997-su], but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww; @Sliva2002-oz]. Styles were difficult to determine for the original researchers [@Rice1994-rk], and I had the same difficulty. In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case. This is indicative of relational identity. There are no environmental reasons to assume differences in projectile points, as each site was in a similar ecozone. Each hunter would have been using the points for the same game and differences in point styles would have been primarily cultural [@Sliva2002-oz].

There is one other aspect to these relations that I believe played an important role. A central way people identify themselves is by gender. Gender roles, like identity, can also vary in complex ways, but for the purposes of this analysis I will use a simplified model. Women made pottery and men made projectile points. This was not always true of course, but this fits the available data for the Hohokam [@Crown1996-xb; @VanPool2010-im, p. 253]. An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7]. On the other hand, projectile points were almost exclusively buried with males. This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.

@@

The network for architecture, ceramics, and projectile points, were created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site. This is a simple measure of similarity that does not take into account abundance. It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.

```{r sites, echo=FALSE}
tbls = tbls + 1
cap = "Project Sites and Data"
sites %>% 
  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
  # as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>% 
  # theme_article() %>% 
  # set_caption(cap)
  knitr::kable(caption= cap,
               booktabs = T,
               format.args = list(big.mark = ",",
                                  scientific = FALSE)) %>%
  kable_styling(latex_options = "scale_down")
```

The types of similarity used in this study are varied depending on the type of network. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj], but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.

I used ranked links to keep consistent values between networks. This involved calculating the strength of similarity between each site in the network and then ranking the strength of similarity. The top **n** connections between each site were kept and the rest were discarded. In practice, not every node will have the same number of connections. Some may have fewer ties if there are not enough nodes that are similar. More ties can exist when there are ties in the ranks. Because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top **n** of multiple other nodes. In the latter case, this is a good indication of a central node in the network. Because I am using several types of networks, there is no common cutoff value I can use for the number of ties to keep. Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top 3-10 ties. Meaning that each metric is calculated for a network composed of ties with the top 3 connections, and the process is repeated for new networks composed of ties with the top 4 connections, and so forth. Examining the range of these metrics provides a more robust analysis.

Two network metrics are discussed in this analysis: eigenvector centrality and multilayer network correlation--the second will be discussed in the next section. Node centrality is a common way to quantify networks [@Borgatti2005-iz]. Centrality is a measure of the influence of the node on the network. Nodes with higher centrality generally derive or generate greater benefits from the network. For example, if I have many friends, then I can call in more favors. Eigenvector centrality is one way to measure node centrality and is commonly used in archaeology. This metric describes how well a node is connected to the network as a whole and is helpful in comparing different networks containing the same nodes. For example, if my friends have many friends, then I will have a greater advantage then someone with an equal number of friends, but whose friends have few friends. Essentially, this is a way of measuring second-order and beyond connections. Eigenvector centrality is also more robust to missing nodes than other measures [@Peeples2017-kl], a major problem in most archaeological studies.

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network). Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results. Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality. In this case the centrality measure is simply the mean of the eigenvector centraliy for each separate layer, as the layers are not interdependent [@Frost2022-lr]. Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers. These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.

The robustness of results were tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value. This provides a baseline that determines how likely it is to obtain a given result by chance.

Certain caveats must be acknowledged. All archaeologists deal with missing data, which can greatly affect the results of analysis. Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik]. Regardless of statistical methods, unaccounted for biases will still produce invalid results [@Bischoff2021-sf]. The most important bias that may affect these findings is chronology. For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq]. Does that invalidate the network analysis? I do not think so. While the networks involve nodes as sites, what the networks are really meant to represent are groups of people. Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation. Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p. 336]. Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location. The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96]. If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on. It is these communities that the network analysis is attempting to capture. Still, evidence indicates that most sites were contemporaneous. For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya]. It is best, though, to think of these interpretations as estimations based on incomplete data.

@@

The simplest network is the spatial network. This network was created by calculating the Euclidean distance from every site to every other site. Euclidean distance is a shortcut that does not represent actual travel routes. Least cost paths provide more accurate data [see @Caseldine2022-uu for an application in Tonto Basin], yet an analysis in similar terrain indicates that for distances longer than those in this study the benefits of least cost path distances in place of Euclidean distance are negligible [@Bischoff2017-gx]. Least cost paths are also hampered by the presence of Roosevelt Lake and other modern impacts on the landscape. The spatial network created here is equivalent to proximal point analysis, which was used in one of the earliest network studies in archaeology [@Terrell1977-co] and continues to be used to study potential pathways of interaction [see @Broodbank2000-pg; @Collar2013-fb]. The inclusion of a spatial network serves as a null hypothesis to test the other networks against.

@@

The architectural components consisted of compounds, roomblocks, and platform mounds. Compounds were the typical residential structures in the area and featured multiple houses built from masonry and/or adobe around a courtyard. Platform mounds were formed from large masonry and adobe retaining walls with a cell-like interior filled with trash and rubble. They were common in the Hohokam region throughout the Classic period. Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku]. Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture. Thus, if a site is labeled a platform mound it may have other architecture. Sites were thus labeled to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.

@@

The ceramic network consists of 55 types--plainware and decorated--ranging from 181,666 Plain Brown Ware sherds to a single Black Mesa Black-on-white sherd. A single sherd is not much, but it is still indicative of trade or movement. The median sherd count was 157 for all types. The ceramic data was accessed from the cyberSW database [@Mills2020-tb], which keeps the full citations for the ceramic data. The standard caveats for ceramic analysis apply to this analysis as well: problems with sherd misidentification and data errors.

@@

The point networks were expected to represent relational identification among men, at least in this case study. Because relational identification is related to frequent interaction, spatial distance is a crucial component. It is much harder to interact with someone when they are far away. Thus, the correlation between the point and spatial networks also makes theoretical sense.
